\Title

### 2019-06-21 manuscript/manuscript.tex
Update reference to Lin2019

To take into account the curvature of the Earth, many global forward modelling methods
are defined in geocentric spherical coordinates.
A common approach is to discretize the Earth into tesseroids \citep{Anderson1976},
i.e.~spherical prisms, which are defined by pairs of latitude, longitude, and radial
boundaries (Fig.~\ref{fig:tesseroid}).
The gravitational fields generated by a
tesseroid on any external point are given by volume
integrals that must be numerically approximated.
The literature offers two main approaches: one involves Taylor series expansion
\citep{Heck2007, Grombein2013} while the other makes use of Gauss-Legendre
Quadrature (GLQ)
\citep{Asgharzadeh2007, Wild-Pfeiffer2008, Li2011, Uieda2016, Lin2018}.
The Taylor series expansion is not well suited to develop an algorithm for
a density varying with depth according to an arbitrary continuous function.
Different series expansion terms would have to obtained for each density function
desired.
Conversely, an arbitrary continuous density function can be included in the GLQ without
any change to the integration method.
Thus, we will focus only on GLQ-based methods henceforth.

@@

Two recent publications present alternative approaches for calculating the gravitational
fields of homogeneous tesseroids and incorporate methodologies for tesseroids with
variable densities in depth.
\citet{Fukushima2018} analytically integrates the volumetric integral for the
gravitational potential in the radial direction, obtaining a surface integral, which is
then numerically solved by conditionally splitting the tesseroid and applying the double
exponential quadrature rule.
The gradient of the potential and the Marussi tensor components are computed by finite
differences.
\citet{Fukushima2018} also generalized their method to tesseroids with a radial
polynomial density function of arbitrary degree.
\citet{Lin2018} compared the different integration and discretization methodologies for
homogeneous tesseroids.
From this analysis they developed a combined method:
for computation points near the tesseroid, they use a GLQ integration with an adaptive
discretization based on \citet{Uieda2016} but only applied to the horizontal dimensions.
If the computation point is farther than a certain truncation distance,
a second order Taylor series approximation is applied instead along with the regular
subdivision developed by \citet{Grombein2013}.
\citet{Lin2018} also introduced a variation of their combined method to compute the
gravitational fields generated by tesseroids with a linearly varying density in the
radial dimension.

Both the \citet{Lin2018} and the \citet{Fukushima2018} studies limit the radial density
variation to polynomial functions.
While most continuous and smooth functions can be approximated by piecewise linear
functions, the choice of a discretization interval is neither straight forward nor
automatic for the general case.
Although many piecewise linear approximation algorithms that automate this process
exist \citep{Ketkov1969, Vandewalle1975, Imamoto2008, Ahmadi2013}, they either require
a fixed number of discretization intervals or are meant to be used only on convex
functions.
Using such algorithms would limit the range of density functions that can be assigned
to the tesseroids and would not guarantee a fully automatic process.
Furthermore, it is well known that the use of high-degree polynomials to approximate a
highly variable function produces unstable results when extrapolating beyond the data
domain.
These shortcomings could make piecewise linear or high-degree polynomial density
functions unwieldy for non-linear gravity inversions \citep[e.g.][]{Uieda2017} if the
density function in depth is highly variable.

We present a new algorithm for computing the gravitational potential and its gradient
generated by a tesseroid with an arbitrary continuous density function on an external
point.
It is based on the three dimensional GLQ integration, a two dimensional version of the
adaptive discretization of \citet{Uieda2016} (following \citet{Lin2018}),
and a new density-based radial discretization algorithm.
To ensure the accuracy of the numerical approximation, we empirically determine
optimal values for the controlling parameters by comparing the numerical results with
analytical solutions for spherical shells.
Finally, we apply the methodology to model the Neuqu\'en basin, Argentina, using
tesseroids with linear and exponentially increasing density with depth.

@@

\citet{Ku1977} noticed that the GLQ integration
becomes less accurate when the computation point is closer to the
mass element.
One way to prevent this from happening would be to increase the GLQ order.
Doing so would uniformly increase the number of point masses inside the
tesseroid volume.
However, an increase in the point mass concentration is only required close to the
computation point \citep{Uieda2016}.
Alternatively, \citet{Li2011} proposed an adaptive
discretization algorithm which keeps the GLQ order fixed and divides the
tesseroid based on a ratio between the distance to the computation
point and its dimensions.
This algorithm produces a more efficient computation because an increased concentration
of point masses is produced where it is needed more.
\citet{Uieda2016} developed a modified version of this algorithm along with an efficient
computational implementation.
Both algorithms perform tesseroid subdivisions in the latitudinal,
longitudinal and radial directions, thus we can define them as three-dimensional
adaptive discretization algorithms.
On the other hand, \citet{Lin2018} proposed a two dimensional discretization algorithm
that subdivides the tesseroid only on the latitudinal and longitudinal directions.
Removing a dimension from the discretization makes the computation more efficient by
reducing the number of tesseroids in the model, while
retaining an acceptable accuracy \citep{Lin2018}.

Here we will follow \citet{Lin2018} and use a two dimensional version of the adaptive
discretization of \citet{Uieda2016}.
What follows is a summary of the algorithm and the reader is referred to
\citet{Uieda2016} for a detailed description.

@@

\citet{Uieda2016} compared the numerical integration of homogeneous density tesseroids
with the analytical solution of a spherical shell \citep{Mikuska2006, Grombein2013} in
order to obtain default values for the distance-size ratio $D$.
We will follow this idea but for our needs the spherical shell must
have the same density function of radius as our tesseroid model.
\citet{Lin2018} show the analytical solution of the gravitational potential generated by
a spherical shell with linear density in the radial coordinate.
Applying the Newton's Shell Theorem \citep{Chandrasekhar1995, Binney2008},
we derive expressions for the gravitational potential of a spherical shell with
linear, exponential, and sinusoidal density functions
(see Appendix~\ref{sec:shell}).

@@

By including the density function into the GLQ integration, the method described here
can be applied without modification to tesseroids with density following any continuous
function of radius.
A density-based discretization algorithm divides the tesseroid in the radial dimension
to ensure accurate integration of the density function.
This algorithm is independent of the GLQ integration and could potentially be used to
determine an optimal discretization when approximating a density function by piecewise
linear \citep{Lin2018} or piecewise polynomial \citep{Fukushima2018} functions.

@@

\noindent The first term on this equation reproduces the potential generated
by a spherical shell with variable density $\rho(r') = ar'$, while the second
term constitutes the potential generated by a spherical shell with homogeneous
density $\rho = b$ \citep{Mikuska2006,Grombein2013}.
Eq~\ref{eq:shell-pot-linear} is in agreement with the one obtained by \citet{Lin2018}.

>>>>

To take into account the curvature of the Earth, many global forward modelling methods
are defined in geocentric spherical coordinates.
A common approach is to discretize the Earth into tesseroids \citep{Anderson1976},
i.e.~spherical prisms, which are defined by pairs of latitude, longitude, and radial
boundaries (Fig.~\ref{fig:tesseroid}).
The gravitational fields generated by a
tesseroid on any external point are given by volume
integrals that must be numerically approximated.
The literature offers two main approaches: one involves Taylor series expansion
\citep{Heck2007, Grombein2013} while the other makes use of Gauss-Legendre
Quadrature (GLQ)
\citep{Asgharzadeh2007, Wild-Pfeiffer2008, Li2011, Uieda2016, Lin2019}.
The Taylor series expansion is not well suited to develop an algorithm for
a density varying with depth according to an arbitrary continuous function.
Different series expansion terms would have to obtained for each density function
desired.
Conversely, an arbitrary continuous density function can be included in the GLQ without
any change to the integration method.
Thus, we will focus only on GLQ-based methods henceforth.

@@

Two recent publications present alternative approaches for calculating the gravitational
fields of homogeneous tesseroids and incorporate methodologies for tesseroids with
variable densities in depth.
\citet{Fukushima2018} analytically integrates the volumetric integral for the
gravitational potential in the radial direction, obtaining a surface integral, which is
then numerically solved by conditionally splitting the tesseroid and applying the double
exponential quadrature rule.
The gradient of the potential and the Marussi tensor components are computed by finite
differences.
\citet{Fukushima2018} also generalized their method to tesseroids with a radial
polynomial density function of arbitrary degree.
\citet{Lin2019} compared the different integration and discretization methodologies for
homogeneous tesseroids.
From this analysis they developed a combined method:
for computation points near the tesseroid, they use a GLQ integration with an adaptive
discretization based on \citet{Uieda2016} but only applied to the horizontal dimensions.
If the computation point is farther than a certain truncation distance,
a second order Taylor series approximation is applied instead along with the regular
subdivision developed by \citet{Grombein2013}.
\citet{Lin2019} also introduced a variation of their combined method to compute the
gravitational fields generated by tesseroids with a linearly varying density in the
radial dimension.

Both the \citet{Lin2019} and the \citet{Fukushima2018} studies limit the radial density
variation to polynomial functions.
While most continuous and smooth functions can be approximated by piecewise linear
functions, the choice of a discretization interval is neither straight forward nor
automatic for the general case.
Although many piecewise linear approximation algorithms that automate this process
exist \citep{Ketkov1969, Vandewalle1975, Imamoto2008, Ahmadi2013}, they either require
a fixed number of discretization intervals or are meant to be used only on convex
functions.
Using such algorithms would limit the range of density functions that can be assigned
to the tesseroids and would not guarantee a fully automatic process.
Furthermore, it is well known that the use of high-degree polynomials to approximate a
highly variable function produces unstable results when extrapolating beyond the data
domain.
These shortcomings could make piecewise linear or high-degree polynomial density
functions unwieldy for non-linear gravity inversions \citep[e.g.][]{Uieda2017} if the
density function in depth is highly variable.

We present a new algorithm for computing the gravitational potential and its gradient
generated by a tesseroid with an arbitrary continuous density function on an external
point.
It is based on the three dimensional GLQ integration, a two dimensional version of the
adaptive discretization of \citet{Uieda2016} (following \citet{Lin2019}),
and a new density-based radial discretization algorithm.
To ensure the accuracy of the numerical approximation, we empirically determine
optimal values for the controlling parameters by comparing the numerical results with
analytical solutions for spherical shells.
Finally, we apply the methodology to model the Neuqu\'en basin, Argentina, using
tesseroids with linear and exponentially increasing density with depth.

@@

\citet{Ku1977} noticed that the GLQ integration
becomes less accurate when the computation point is closer to the
mass element.
One way to prevent this from happening would be to increase the GLQ order.
Doing so would uniformly increase the number of point masses inside the
tesseroid volume.
However, an increase in the point mass concentration is only required close to the
computation point \citep{Uieda2016}.
Alternatively, \citet{Li2011} proposed an adaptive
discretization algorithm which keeps the GLQ order fixed and divides the
tesseroid based on a ratio between the distance to the computation
point and its dimensions.
This algorithm produces a more efficient computation because an increased concentration
of point masses is produced where it is needed more.
\citet{Uieda2016} developed a modified version of this algorithm along with an efficient
computational implementation.
Both algorithms perform tesseroid subdivisions in the latitudinal,
longitudinal and radial directions, thus we can define them as three-dimensional
adaptive discretization algorithms.
On the other hand, \citet{Lin2019} proposed a two dimensional discretization algorithm
that subdivides the tesseroid only on the latitudinal and longitudinal directions.
Removing a dimension from the discretization makes the computation more efficient by
reducing the number of tesseroids in the model, while
retaining an acceptable accuracy \citep{Lin2019}.

Here we will follow \citet{Lin2019} and use a two dimensional version of the adaptive
discretization of \citet{Uieda2016}.
What follows is a summary of the algorithm and the reader is referred to
\citet{Uieda2016} for a detailed description.

@@

\citet{Uieda2016} compared the numerical integration of homogeneous density tesseroids
with the analytical solution of a spherical shell \citep{Mikuska2006, Grombein2013} in
order to obtain default values for the distance-size ratio $D$.
We will follow this idea but for our needs the spherical shell must
have the same density function of radius as our tesseroid model.
\citet{Lin2019} show the analytical solution of the gravitational potential generated by
a spherical shell with linear density in the radial coordinate.
Applying the Newton's Shell Theorem \citep{Chandrasekhar1995, Binney2008},
we derive expressions for the gravitational potential of a spherical shell with
linear, exponential, and sinusoidal density functions
(see Appendix~\ref{sec:shell}).

@@

By including the density function into the GLQ integration, the method described here
can be applied without modification to tesseroids with density following any continuous
function of radius.
A density-based discretization algorithm divides the tesseroid in the radial dimension
to ensure accurate integration of the density function.
This algorithm is independent of the GLQ integration and could potentially be used to
determine an optimal discretization when approximating a density function by piecewise
linear \citep{Lin2019} or piecewise polynomial \citep{Fukushima2018} functions.

@@

\noindent The first term on this equation reproduces the potential generated
by a spherical shell with variable density $\rho(r') = ar'$, while the second
term constitutes the potential generated by a spherical shell with homogeneous
density $\rho = b$ \citep{Mikuska2006,Grombein2013}.
Eq~\ref{eq:shell-pot-linear} is in agreement with the one obtained by \citet{Lin2019}.
